\ttitle

### 2015-08-22 Chapters/Introduction.tex
Review 1
- Every sentence on a new line
- Fits my OCD
Review 1 - Introduction
- Fixed nearly all incomplete sections
- All todos sorted
- Multiple tables amended for formatting
- More references for jaundice, immunity
- Added table for POSSUM
- Fixed several overfull box issues with TOC

\section{Pancreatic Neoplasia}
%This is the only paper I need for this entire section - pancreatic_section_of_the_british_society_of_gastroenterology_pancreatic_society_of_great_britain_and_ireland_association_of_upper_gastrointestinal_surgeons_of_great_britain_and_ireland_royal_college_of_pathologists_special_interest_group_for_gastro-i_guidelines_2005

@@

%10/08/2015 - Review completete to here

@@

Reconstruction was performed as follows: either a pancreatico-jejunostomy was fashioned using 4-0 Biosyn sutures in a two-layer duct-to-mucosa technique or a pancreatico-gastrostomy was created using 3/0 Biosyn sutures placed in a similar manner. Hepatico-jejunostomy was performed using interrupted 4/0 Biosyn sutures while the gastro-jejunostomy or duodenojejunostomy (in PPPD) was performed using continuous 3/0 Polydioxanone (PDS) sutures in 2-layers. One or two surgical drains were placed and the abdomen was closed after ensuring haemostasis.

\subsection{Postoperative care}
All patients were routinely admitted to the Surgical High Dependency Unit unless intra-operative events necessitated admission to the Intensive Care Unit. A standardised regimen of intravenous fluids, naso-jejunal feeding, mobilisation and physiotherapy was implemented in all patients. Standard physiological parameters including haemodynamic parameters, renal function and arterial blood gases were used to monitor adequate end organ perfusion. All patients received proton pump inhibitors and Octreotide. Patients were discharged to the general surgical ward as early as possible. Amylase content in the drain fluid was measured on the third post-opertive day to aid in the diagnosis of post-operative pancreatic fistula. Surgical drains and naso-gastric/naso-jeujunal tubes were removed at surgeon's discretion. Patients were discharged when medically fit.

@@

\section{Adjuvant and Neoadjuvant treatment}

@@

\section{Comorbidity and Risk Stratification}
\label{sec:comorbidity_risk_stratification}
\todo{Comorbidity and Risk Stratification - needs major tidying up - will take several days}
Selection of patients suitable for potentially curative surgery not only depends on the resectability criteria that were discussed in section \ref{sec:resectability_criteria} but also on patient fitness. Major pancreatic surgery is associated with significant morbidity. Subjecting patients with poor physiological reserve to a pancreaticoduodenectomy is fraught with risk. It can result in a worse outcome for the patient due to either post-operative mortality or prolonged hospitalisation due to complications resulting in poor quality of life with little difference in longevity as compared to palliation. Patient factors that affect postoperative recovery and long-term survival are therefore important in determining suitability for surgery.

@@

\subsection{Risk Stratification}
\todo{Intro - risk stratification}
%Include other measures - discuss POSSUM, APACHE, mGPS, EPASS and everything else here.
Physiological fitness or reserve may be defined as the ability of the patient's organ systems to respond appropriately and adequately to the stress of surgery. Major surgery has a significant physiological impact on multiple organ systems, especially the cardiorespiratory system. The ability of the cardiorespiratory system as well as other physiological systems including renal, gastrointestinal, hepatic, coagulatory and immunological systems to cope with major surgery and its sequelae plays a major role in determining postoperative outcomes. Therefore, accurate assessment of physiological fitness is important in selecting patients who will benefit from surgery, advising patients of the risks of surgery as it applies to them, preoperative optimisation as well as in perioperative management decisions. A thorough clinical history of co-existing illnesses, general examination and review of patient's medications is followed by more specialist testing to assess fitness. Numerous methods including individual tests and composite scores derived from physiological and biochemical variables have been used for risk assessment.

\subsection{Static Versus Dynamic Testing}
\todo{Static Versus Dynamic Testing}

Conventional tests of cardiac and respiratory function assess either performance at rest or only one component of the cardiorespiratory system. These tests include electrocardiography, echocardiography, exercise tolerance testing, myocardial perfusion scan, pulmonary function tests including spirometry. However, none of these tests adequately measure the ability of the cardiopulmonary and circulatory systems to deliver oxygen to the tissues at times of increased demand. Patients with subclinical cardiopulmonary dysfunction or limitation cannot be identified on these tests alone and it is this group of patients who are most likely to be at greater risk of complications and who will most benefit from targeted prehabilitation. Cardiopulmonary exercise testing overcomes the limitations of the conventional static tests by providing a global assessment of the patient's oxygen delivery mechanisms at times of increased physiological stress and demand.

One of the earliest attempts to develop a dynamic test in surgical patients was reported by Starr and co-workers in 1945 who tried to devise a test that could assess respiratory function and oxygen utilisation in surgical patients before, immediately after and a week after surgery. They reported on the difficulties in obtaining reproducible results but suggested that a slow return to baseline of heart rate, respiratory rate and oxygen utilisation after surgery may indicate patients who are slow to recover. \parencite{starr_convalescence_1945}

\section{Cardiopulmonary Exercise Testing}
%Not sure I need these two sections - delete once done vacillating
\todo{A brief history of CPET in general}
\todo{The physiological basis of CPET}

\subsection{Origins and physiological basis of cardiopulmonary exercise testing}
The measurement of aerobic fitness has been an important aspect of improving athletic performance in competitive sports. Cardiopulmonary exercise testing has been used for a long time for detailed measurement of aerobic fitness in athletes. It has also been used to design customised training regimens aimed at improving performance at activities that place a significant stress on the cardiovascular and respiratory systems by allowing the optimisation of the individual components of the oxygen delivery system.

\subsection{Cardiopulmonary Exercise Test Methodology}
\label{sec:cpx_method}

\begin{figure}[htbp]
	\centering
	\includegraphics[width=\textwidth]{Figures/cpet-zan}
	\caption{ZAN-600 Cycle Erogmeter and CPET suite. (www.nspirehealth.com)}
	\label{fig:cpet-zan}
\end{figure}

CPET is composed of several components that involve measuring not only the response of the cardiac and respiratory system to exercise but the test also helps establish the adequacy of this response to sustain oxygen delivery to skeletal muscle as demand increases with increasing exercise.

Cardiopulmonary exercise tests were performed in the Department of Respiratory Medicine at the Glasgow Royal Infirmary using the ZAN-600 CPET suite (nSpire Health, Longmont, CO 80501, USA). The equipment was calibrated regularly to the standards set by the manufacturer and currently published guidelines[]. All tests were performed by specialist respiratory physiologists. Suitable equipment for cardiopulmonary resuscitation were available in the department in case of unexpected problems. The department was situated within the main hospital premises and therefore was easily accessible to the hospital cardiac arrest team. All patients were fully informed of the steps involved in the procedure, the reasons for performing the test as well as the risks involved.

Spirometry was performed in all patients prior to CPET. Capillary blood gases were measured in all patients after CPET. An electronically braked cycle ergometer was used to increase resistance to pedalling in preset increments. A tight-fitting face mask was placed on the patient covering the nose and the mouth. This allowed breath-by-breath gas analysis thus allowing measurement of several respiratory parameters as listed in table []. 12-lead electrocardiogram was recorded at the same time.

The test started with an initial 3-minute rest period to allow measurement of baseline parameters. This was followed by an incremental work-load test that involved the patient pedalling approximately at 60 revolutions per minute while the resistance to pedalling was gradually increased in preset increments. The test was terminated when patients reached volitional fatigue (maximal exercise tolerance), significant ischaemic changes on ECG or for other safety reasons.

The parameters measured at spirometry are shown in Table \ref{table:spirometry} and those measured during cardiopulmonary exercise testing are shown in Table \ref{table:cpet_parameters} on p\pageref{table:cpet_parameters}.

\subsection{Measuring the Anaerobic Threshold}

\begin{figure}[htbp]
	\centering
	\includegraphics[width=\textwidth]{Figures/cpet_9panel.pdf}
	\caption{9-panel view of trending parameters during incremental cardiopulmonary exercise.}
	\label{fig:cpet_9panel}
\end{figure}

The anaerobic threshold (variously described as the lactate threshold or ventilatory threshold) is the point during exercise when oxygen demand by exercising skeletal muscle outstrips supply. Therefore, muscle tissues use anaerobic respiration to supplement aerobic respiration to continue generation of ATP. The resulting metabolic lactic acidosis is almost immediately compensated by the bicarbonate buffer as below: 
\begin{equation} \label{eq:bicarb_buffer}
	H^+ + HCO3^- \Longleftrightarrow H_2CO_3 \Longleftrightarrow H_2O + CO_2
\end{equation}
The resulting excess $CO_2$ is exhaled and is one of the many parameters measured during cardiopulmonary exercise testing. This transition from aerobic to anaerobic respiration may be determined using the V-slope method \parencite{sue_metabolic_1988} or the ventilatory equivalents method \parencite{beaver_new_1986}. Most centres, like ours, use both methods supplemented by information from a variety of other parameters to enable accurate determination of the anaerobic threshold as recommended by the American Thoracic Society/American College of Chest Physicians Statement on cardiopulmonary exercise testing \parencite{society_ats/accp_2003}.

@@

During aerobic phase of exercise, $\dot{V}_{O_2}$ and $\dot{V}_{CO_2}$ share a linear relationship as shown in the left half of Figure \ref{fig:cpet_vslope} on p\pageref{fig:cpet_vslope}. However, as anaerobic respiration starts to supplement aerobic respiration, $\dot{V}_{CO_2}$ increases disproportionately to $\dot{V}_{O_2}$ as a direct result of the respiratory buffer described in equation \ref{eq:bicarb_buffer} on p\pageref{eq:bicarb_buffer}. This results in a distinct difference in the slope of the relationship between $\dot{V}_{O_2}$ and $\dot{V}_{CO_2}$ during aerobic exercise (Line A) versus during anaerobic exercise (Line B). The point at which the two slopes intersect is the anaerobic threshold and the $\dot{V}_{O_2}$ at this point in exercise is commonly referred to as the anaerobic threshold, $\dot{V}_{O_2}$AT or simply AT.

@@

\section{Description of CPET Parameters}
\label{sec:cpx_parameters}
\input{Tables/c1_cpet_parameters}

\subsection{Exercise Load}
The most common form of cardiopulmonary exercise testing for clinical use involves a cycle ergometer with steadily increasing resistance delivered through electric braking. This allows accurate measurement of work load in Watts. The relationship between $\dot{V}_{O_2}$ and work rate is usually linear and the slope of this relationship is independent of sex, age or height. An abnormality in this relationship is usually due to cardiopulmonary or circulatory causes.

\subsection{Minute Ventilation, $\dot{V}_E$}
Minute ventilation or respiratory minute volume is the volume of air that is inhaled/expired in a minute.
\begin{equation} \label{eq:VE=VTxBf}
	\dot{V}_E = \dot{V}_T \times Bf
\end{equation}
where $\dot{V}_T$ = Tidal Volume and $Bf$ = Breathing Frequency.

@@

\subsection{Oxygen Uptake, $\dot{V}_{O_2}$}

@@

\subsection{Oxygen Pulse, $O_2Pulse$}
Oxygen pulse is defined as the oxygen uptake per heart beat.
\begin{equation} \label{eq:oxygen_pulse}
	O_2Pulse = \frac{\dot{V}_{O_2}}{Heart\ rate}
\end{equation}
While some authors have suggested that oxygen pulse may be a surrogate for stroke volume others disagree. The clinical application of oxygen pulse in surgical patients remains unclear.

\subsection{Respiratory Exchange Ratio, RER}
The ratio of $\dot{V}_{CO_2}/\dot{V}_{O_2}$ is called the Respiratory Exchange Ratio. An RER greater that 1.0 may be caused either by lactic acidosis or due to hyperventilation. The RER is also a marker of the fuel being used for metabolism with RER less than 1.0 indicating mixed fuel source in the form of carbohydrate and fat while an RER of 1.0 or greater indicates a primarily carbohydrate source.

\subsection{Ventilatory Equivalent for $O_2$ and $CO_2$, $\dot{V}_E/\dot{V}_{O_2}$, $\dot{V}_E/\dot{V}_{CO_2}$}
The changes in $\dot{V}_E/\dot{V}_{O_2}$ and $\dot{V}_E/\dot{V}_{CO_2}$ during exercise provide valuable information regarding the ventilatory response to exercise. Both $\dot{V}_E/\dot{V}_{O_2}$ and $\dot{V}_E/\dot{V}_{CO_2}$ tend to decrease initially during exercise. However, as the anaerobic threshold is passed, $\dot{V}_E/\dot{V}_{O_2}$ starts increasing before $\dot{V}_E/\dot{V}_{CO_2}$. This change in direction is yet another method to confirm the anaerobic threshold. $\dot{V}_E/\dot{V}_{CO_2}$ eventually starts increasing as respiratory compensation of metabolic acidosis results in increased $\dot{V}_E$.

\subsection{End-tidal $O_2$ and $CO_2$, $P_{ET_{O_2}}$, $P{ET_{CO_2}}$}
$PET_{O_2}$ and $PET_{CO_2}$ are the partial pressures of $O_2$ and $CO_2$ at the end of an exhaled breath and are closely related to $P_aO_2$ and $P_aCO_2$ respectively. $P{ET_{CO_2}}$ is dependent on pulmonary gas-exchange which is in turn influenced by the right ventricular output, pulmonary blood flow and alveolar gas exchange. The changes in $P_{ET_{O_2}}$ and $P{ET_{CO_2}}$ during exercise help identify ventilation-perfusion mismatch as well as hyperventilation.

\subsection{Heart Rate, HR}
\label{sec:heart_rate}
The heart rate response during exercise in healthy individuals is a linear function of $\dot{V}_{O_2}$ increasing linearly with increasing work load and increasing $\dot{V}_{O_2}$. The difference between the predicted peak heart rate and the observed peak heart rate is called the Heart Rate Reserve or HRR. Failure to achieve the predicted peak heart rate or a wide HRR may be due to cardiac disease or due to medication used to treat cardiovascular disorders such as beta-blockers or calcium-channel blockers. This information in conjunction with 12-lead ECG evidence of ischaemia provides undeniable evidence of primary cardiac dysfunction.

\subsection{Breathing frequency, $B_f$}
In healthy individuals, breathing frequency is one of the most important respiratory responses to exercise in order to increase oxygen uptake. However, a variety of respiratory disorders including chronic obstructive pulmonary disease or restrictive disorders such as interstitial lung disease can impair this response. Morbidly obese patients may also have additional restrictions on respiratory response due to their body habitus. On the other hand, hyperventilation in an otherwise healthy patient may make interpretation of the results difficult although the analysis of trends in other parameters such as $P_{ET_{O_2}}$ and $P{ET_{CO_2}}$ aids in the diagnosis of hyperventilation versus true respiratory problems.

\section{Role of CPET in preoperative assessment}
\label{sec:cpx_roleInAssessment}
\subsection{General Surgery}

>>>>

\section{Pancreatic neoplasia}
%This is the only paper I need for this entire section - pancreatic_section_of_the_british_society_of_gastroenterology_pancreatic_society_of_great_britain_and_ireland_association_of_upper_gastrointestinal_surgeons_of_great_britain_and_ireland_royal_college_of_pathologists_special_interest_group_for_gastro-i_guidelines_2005

@@

@@

Reconstruction was performed as follows: either a pancreatico-jejunostomy was fashioned using 4-0 Biosyn sutures in a two-layer duct-to-mucosa technique or a pancreatico-gastrostomy was created using 3/0 Biosyn sutures placed in a similar manner. 
Hepatico-jejunostomy was performed using interrupted 4/0 Biosyn sutures.
Gstro-jejunostomy or duodenojejunostomy (in PPPD) was performed using continuous 3/0 Polydioxanone (PDS) sutures in 2-layers. 
One or two surgical drains were placed and the abdomen was closed after ensuring haemostasis.

\subsection{Postoperative care}
All patients were routinely admitted to the Surgical High Dependency Unit unless intra-operative events necessitated admission to the Intensive Care Unit. 
A standardised regimen of intravenous fluids, naso-jejunal feeding, mobilisation and physiotherapy was implemented in all patients. 
Standard physiological parameters including haemodynamic measurements, renal function and arterial blood gases were used to monitor adequate end organ perfusion. 
All patients received proton pump inhibitors and Octreotide. 
Patients were discharged to the general surgical ward as early as possible. 
Amylase content in the drain fluid was measured on the third post-operative day to aid in the diagnosis of post-operative pancreatic fistula. 
Surgical drains and naso-gastric/naso-jeujunal tubes were removed at the surgeon's discretion. 
Patients were discharged when medically fit.

@@

\section{Adjuvant and neoadjuvant treatment}

@@

\section{Preoperative risk stratification}
\label{sec:comorbidity_risk_stratification}

Selection of patients suitable for potentially curative surgery not only depends on the resectability criteria that were discussed in section \ref{sec:resectability_criteria} but also on patient fitness \parencite{bilimoria_national_2007, sandroussi_sociodemographics_2010}. 
Major pancreatic surgery is associated with significant morbidity. 
Subjecting patients with poor physiological reserve to a pancreaticoduodenectomy is fraught with risk. 
It can result in a worse outcome for the patient due to either post-operative mortality or prolonged hospitalisation due to complications resulting in poor quality of life with little difference in survival when compared to palliation. 
Patient factors that affect postoperative recovery and long-term survival are therefore important in determining suitability for surgery.

@@

\subsection{Risk stratification}
Physiological fitness or reserve may be defined as the ability of the patient's organ systems to respond appropriately and adequately to the stress of surgery. 
Major surgery has a significant physiological impact on multiple organ systems, especially the cardiorespiratory system. 
The ability of the cardiorespiratory system as well as other organ systems including renal, gastrointestinal, hepatic, coagulatory and immunological systems to cope with major surgery and its sequelae plays a major role in determining postoperative outcomes. 
Therefore, accurate assessment of physiological fitness is important in selecting patients who will benefit from surgery, advising patients of the risks of surgery as it applies to them, preoperative optimisation as well as in perioperative management decisions. 
A thorough clinical history of co-existing illnesses, general examination and review of patient's medications is followed by more specialist testing to assess fitness. 
Numerous methods including individual tests and composite scores derived from physiological and biochemical variables have been used for risk assessment.

The Physiological and Operative Severity Score for the enumeration of Mortality and Morbidity (POSSUM) is one of the most used scores for risk-prediction in general surgery \parencite{copeland_possum:_1991}.
It was devised by Copeland and co-workers as a method of predicting risk-adjusted mortality and morbidity rates for comparative audit between different surgical departments \parencite{copeland_comparative_1993} and individual surgeons \parencite{copeland_risk-adjusted_1995}.
While some authors have reported it to be useful in predicting postoperative outcomes after pancreatic surgery \parencite{pratt_possum_2008} others have found that POSSUM was not a good risk prediction model in this group of patients \parencite{khan_evaluation_2003, kocher_risk-adjustment_2005, tamijmarane_application_2008, de_castro_evaluation_2009}.
The physiology component of the POSSUM score (POSSUM Physiology Score, PPS) which takes into account both comorbidity and specific preoperative biochemical parameters was used as an objective static measure of comorbidity in this thesis. 
The POSSUM Physiology Score has been reported to be associated with preoperative systemic inflammation and cancer-specific survival in patients who underwent potentially curative surgery for colorectal cancer \parencite{richards_relationship_2010}.
The variables used in the calculation of the POSSUM Physiology Score are shown in Table \ref{table:intro_possum} on p\pageref{table:intro_possum}.
As the Glasgow Coma Scale in all patients undergoing pancreaticoduodenectomy was 15, this variable was excluded in the calculation of the score in this thesis.

\input{Tables/c1_possum}

More recently, elevated preoperative systemic inflammation as measured by the modified Glasgow Prognostic Score (mGPS) has been reported to be associated with increased incidence of infective complications after colorectal surgery \parencite{moyes_preoperative_2009} as well as oesophageal resections \parencite{vashist_glasgow_2010}.

POSSUM and other similar composite scores such as the APACHE II (Acute Physiology and Chronic Health Evaluation II) \parencite{knaus_apache_1985} and Estimation of Physiologic Ability and Surgical Stress (E-PASS)  \parencite{haga_estimation_1999, hashimoto_is_2010} have not found widespread use for risk prediction in pancreatic surgery.
While these scores are either useful for comparative audit between patient populations or in the case of APACHE II, for monitoring response to therapy in intensive care, they are of limited value when applied to individual patients.

\subsection{Static versus dynamic testing}

Conventional tests of cardiac and respiratory function assess either performance at rest or only one component of the cardiorespiratory system. 
These tests include electrocardiography, echocardiography, exercise tolerance testing, myocardial perfusion scan, pulmonary function tests including spirometry. 
However, none of these tests adequately measure the ability of the cardiopulmonary and circulatory systems to deliver oxygen to the tissues at times of increased demand. 
Patients with sub-clinical cardiopulmonary dysfunction or limitation cannot be identified on these tests alone. 
It is this group of patients who are most likely to be at greater risk of complications and who will most benefit from targeted prehabilitation. 
Moreover, composite score such as POSSUM do not provide information on the ability of the patient to cope with physiological stress or on how their risk may be mitigated before surgery.
Cardiopulmonary exercise testing overcomes the limitations of the conventional static tests by providing a global assessment of the patient's oxygen delivery mechanisms at times of increased physiological demand.

\subsection{Pancreas-specific risk factors}
In addition to all of the above risk factors, post pancreatectomy complications are significantly associated with factors unique to pancreatic disease.
Postoperative pancreatic fistula is one of the most dreaded complications after a pancreaticoduodenectomy and is discussed in more detail in Section \ref{sec:ch_intro_POPF}.
Soft pancreas texture, narrow pancreatic duct, pathology that does not obstruct the pancreatic duct such as ampullary or duodenal cancers are independent risk factors for pancreatic fistula \parencite{pratt_risk_2008}.
These organ-specific risk factors are independent of comorbidity and patient fitness \parencite{deoliveira_assessment_2006}.
Lin and co-workers reported that in a study of 1891 patients undergoing pancreaticoduodenectomy, only soft pancreatic gland texture and coronary disease were related to postoperative pancreatic fistula on multivariate analysis \parencite{lin_risk_2004}.
These pancreas-specific risk factors are not modifiable before surgery, have been studied extensively before and are not the subject of this thesis.

\section{Cardiopulmonary exercise testing}

\subsection{Origins of preoperative cardiopulmonary exercise testing}

The idea of measuring $\dot{V}_{O_2}$ (oxygen consumption) and v$\dot{V}_{CO_2}$ (expired carbon-dioxide) as a measure of cardiopulmonary response to exercise was described in the early twentieth century. 
In 1924, Hill and co-workers reported on adaptations made to an existing apparatus to allow measurement of oxygen consumption during exercise \parencite{hill_muscular_1924}. 
Their method involved the subject running while connected to a large bag slung on their back to collect expired gases. 
The subjects then returned to the laboratory where they were connected to a fairly large apparatus involving several more large bags and other equipment that filled the most part of a large room. 
But the premise of their method remains unchanged to this date; to examine the response of the cardiopulmonary system to exercise.
Such methods were refined by other physiologists to reduce the chance of errors and to devise more practical methods suitable for medical use \parencite{katz_ln_metabolic_1934, sutton_estimation_1940}.

One of the earliest attempts to apply these dynamic tests in surgical patients was reported by Starr and co-workers in 1945 \parencite{starr_convalescence_1945}.
They tried to devise a test that could assess respiratory function and oxygen utilisation in surgical patients before, immediately after and a week after surgery. 
They reported on the difficulties in obtaining reproducible results but suggested that a slow return to baseline of heart rate, respiratory rate and oxygen utilisation after surgery could identify patients who are slow to recover.

However, cardiopulmonary exercise testing in its current form where breath-by-breath analysis of gas exchange is combined with real-time electrocardiography during incremental exercise was devised and popularised by Karlman Wasserman and his team of physiologists.
Wassermann and his team reported on the ventilatory response to exercise and how it may be used to identify cardiorespiratory disease in 1964 \parencite{naimark_continuous_1964}.
The pioneering work on the detection of anaerobic threshold was also done by his team and applied in the assessment of patients with cardiac disease \parencite{wasserman_detecting_1964, wasserman_lactate_1967, wasserma.k_anaerobic_1973}.

The use of cardiopulmonary exercise testing as part of preoperative assessment was first reported by Older and co-workers.
They conducted a prospective study of 187 patients over the age of 60 undergoing major abdominal surgery  \parencite{older_preoperative_1993}. 
All patients underwent a symptom-limited exercise test on a cycle ergometer with real-time 12-lead ECG monitoring. 
The average $\dot{V}_{O_2}$ across all patients was 12.4 ml/min/kg. 
There were a total of 11 deaths related to cardiovascular causes (5.9\%) and 3 deaths due to non-cardiovascular causes. 
They reported that 10 out of the 11 deaths due to cardiovascular causes occurred in patients with $\dot{V}_{O_2}AT<11$ml/kg/min (n=55).
Only one patient in the group with $\dot{V}_{O_2}AT\geq11$ml/kg/min (n=132) died of cardiovascular complications (18\% vs 0.8\%, p$<$0.001). 
Eight out of the 10 patients in the low $\dot{V}_{O_2}$AT group who died also had evidence of myocardial ischaemia (p$<$0.01).

\subsection{Cardiopulmonary exercise test methodology}
\label{sec:cpx_method}

\begin{figure}[htbp]
	\centering
	\includegraphics[width=\textwidth]{Figures/cpet-zan}
	\caption{ZAN-600 Cycle Erogmeter and CPET suite.}
	(Image from www.nspirehealth.com)
	\label{fig:cpet-zan}
\end{figure}

Cardiopulmonary exercise tests were performed in the Department of Respiratory Medicine at the Glasgow Royal Infirmary using the ZAN-600 CPET suite (nSpire Health, Longmont, CO 80501, USA). 
All tests were performed by specialist respiratory physiologists. 
Suitable equipment for cardiopulmonary resuscitation were available in the department in case of unexpected problems. 
The department was situated within the main hospital premises and therefore was easily accessible to the hospital cardiac arrest team. 
All patients were fully informed of the steps involved in the procedure, the reasons for performing the test as well as the risks involved.

Spirometry was performed in all patients prior to cardiopulmonary exercise testing. 
Capillary blood gases were measured in all patients after the test. 
The gas and flow analysis equipment were calibrated prior to every test.
An electronically braked cycle ergometer was used to increase resistance to pedalling in preset increments. 
A tight-fitting face mask was placed on the patient covering the nose and the mouth. 
Breath-by-breath gas analysis and 12-lead electrocardiography were displayed in real-time to the investigator allowing measurement of several cardiac and respiratory parameters as they changed during exercise.

The test started with an initial 3-minute rest period to allow measurement of baseline parameters. 
This was followed by an incremental work-load test that involved the patient pedalling at approximately 60 revolutions per minute while the resistance to pedalling was gradually increased in preset increments. 
The test was terminated when patients reached volitional fatigue (maximal exercise tolerance), significant ischaemic changes on ECG or for other safety reasons.

The oxygen consumption at this point (considered `peak' exercise capacity) is called $\dot{V}_{O_2}$Peak.
This is different from $\dot{V}_{O_2}$Max which is identified by the plateauing of the increasing $\dot{V}_{O_2}$ when the physiological limits of oxygen consumption are met in an individual.
$\dot{V}_{O_2}$Max is only observed in very fit individuals or well-trained athletes.
This seldom happens in surgical patients and $\dot{V}_{O_2}$Peak is used as a surrogate for $\dot{V}_{O_2}$Max.

The list of parameters measured is provided in Tables \ref{table:spirometry} and \ref{table:cpet_parameters} on p\pageref{table:spirometry}.
A detailed description of these parameters is provided in Section \ref{sec:cpx_parameters}.

\subsection{Measuring the anaerobic threshold}
\label{sec:cpx_at_method}

\begin{figure}[htbp]
	\centering
	\includegraphics[width=\textwidth]{Figures/cpet_9panel.pdf}
	\caption{9-panel view of trending parameters during cardiopulmonary exercise.}
	\label{fig:cpet_9panel}
\end{figure}

The anaerobic threshold (variously described as the lactate threshold or ventilatory threshold) is the point during exercise when oxygen demand by exercising skeletal muscle outstrips supply. 
Therefore, muscle tissues use anaerobic respiration to supplement aerobic respiration to continue generation of ATP. 
The resulting metabolic lactic acidosis is almost immediately compensated by the bicarbonate buffer as below: 
\begin{equation} \label{eq:bicarb_buffer}
	H^+ + HCO3^- \Longleftrightarrow H_2CO_3 \Longleftrightarrow H_2O + CO_2
\end{equation}
The resulting excess $CO_2$ is exhaled and is one of many parameters measured during cardiopulmonary exercise testing. 
This transition from aerobic to anaerobic respiration may be determined using the V-slope method \parencite{sue_metabolic_1988} or the ventilatory equivalents method \parencite{beaver_new_1986}. 
Most centres, like ours, use both methods supplemented by information from a variety of other parameters to enable accurate determination of the anaerobic threshold.
This approach is also recommended by the American Thoracic Society/American College of Chest Physicians Statement on cardiopulmonary exercise testing \parencite{society_ats/accp_2003}.

@@

During aerobic phase of exercise, $\dot{V}_{O_2}$ and $\dot{V}_{CO_2}$ share a linear relationship as shown in the left half of Figure \ref{fig:cpet_vslope} on p\pageref{fig:cpet_vslope}. 
However, as anaerobic respiration starts to supplement aerobic respiration, $\dot{V}_{CO_2}$ increases disproportionate to $\dot{V}_{O_2}$ as a direct result of the respiratory buffer described in equation \ref{eq:bicarb_buffer} on p\pageref{eq:bicarb_buffer}. 
This results in a distinct difference in the slope of the relationship between $\dot{V}_{O_2}$ and $\dot{V}_{CO_2}$ during aerobic exercise (Line A) versus during anaerobic exercise (Line B). 
The point at which the two slopes intersect is the anaerobic threshold and the $\dot{V}_{O_2}$ at this point in exercise is commonly referred to as the anaerobic threshold or $\dot{V}_{O_2}$AT.

@@

\section{Description of CPET parameters}
\label{sec:cpx_parameters}
\input{Tables/c1_cpet_parameters}

\subsection{Exercise load}
The most common form of cardiopulmonary exercise testing for clinical use involves a cycle ergometer with steadily increasing resistance delivered through electric braking. 
This allows accurate measurement of work load in Watts. 
The relationship between $\dot{V}_{O_2}$ and work rate is usually linear and the slope of this relationship is independent of sex, age or height. 
An abnormality in this relationship is usually due to cardiopulmonary or circulatory causes.

\subsection{Minute ventilation}
Minute ventilation or respiratory minute volume ($\dot{V}_E$) is the volume of air that is inhaled/expired in a minute.
\begin{equation} \label{eq:VE=VTxBf}
	\dot{V}_E = \dot{V}_T \times Bf
\end{equation}
where $\dot{V}_T$ = Tidal Volume and $Bf$ = Breathing Frequency.

@@

\subsection{Oxygen uptake}

@@

\subsection{Oxygen pulse}
$O_2Pulse$ is defined as the oxygen uptake per heart beat.
\begin{equation} \label{eq:oxygen_pulse}
	O_2Pulse = \frac{\dot{V}_{O_2}}{Heart\ rate}
\end{equation}
While some authors have suggested that oxygen pulse may be a surrogate for stroke volume others disagree. 
The clinical application of oxygen pulse in surgical patients remains unclear.

\subsection{Respiratory exchange ratio}
The ratio of $\dot{V}_{CO_2}/\dot{V}_{O_2}$ is called the Respiratory Exchange Ratio (RER). 
An RER greater that 1.0 may be caused either by lactic acidosis or due to hyperventilation. 
The RER is also a marker of the fuel being used for metabolism with RER less than 1.0 indicating mixed fuel source in the form of carbohydrate and fat while an RER of 1.0 or greater indicates a primarily carbohydrate source.

\subsection{Ventilatory equivalents}
The changes in the ventilatory equivalents for $O_2$ ($\dot{V}_E/\dot{V}_{O_2}$) and $CO_2$ ($\dot{V}_E/\dot{V}_{CO_2}$) during exercise provide valuable information regarding the ventilatory response to exercise. 
Both $\dot{V}_E/\dot{V}_{O_2}$ and $\dot{V}_E/\dot{V}_{CO_2}$ tend to decrease initially during exercise. 
However, as the anaerobic threshold is passed, $\dot{V}_E/\dot{V}_{O_2}$ starts increasing before $\dot{V}_E/\dot{V}_{CO_2}$. 
This change in direction is yet another method to confirm the anaerobic threshold. $\dot{V}_E/\dot{V}_{CO_2}$ eventually starts increasing as respiratory compensation of metabolic acidosis results in increased $\dot{V}_E$.

\subsection{End-tidal oxygen and CO2} 
$PET_{O_2}$ and $PET_{CO_2}$ are the partial pressures of $O_2$ and $CO_2$ at the end of an exhaled breath and are closely related to $P_aO_2$ and $P_aCO_2$ respectively. $P{ET_{CO_2}}$ is dependent on pulmonary gas-exchange which is in turn influenced by the right ventricular output, pulmonary blood flow and alveolar gas exchange. 
The changes in $P_{ET_{O_2}}$ and $P{ET_{CO_2}}$ during exercise help identify ventilation-perfusion mismatch as well as hyperventilation.

\subsection{Heart rate}
\label{sec:heart_rate}
The heart rate response during exercise in healthy individuals is a linear function of $\dot{V}_{O_2}$ increasing linearly with increasing work load and increasing $\dot{V}_{O_2}$. 
The difference between the predicted peak heart rate and the observed peak heart rate is called the Heart Rate Reserve or HRR. 
Failure to achieve the predicted peak heart rate or a wide HRR may be due to cardiac disease or due to medication used to treat cardiovascular disorders such as beta-blockers or calcium-channel blockers. 
This information in conjunction with 12-lead ECG evidence of ischaemia provides undeniable evidence of primary cardiac dysfunction.

\subsection{Breathing frequency}
In healthy individuals, breathing frequency ($B_f$) is one of the most important respiratory responses to exercise in order to increase oxygen uptake. 
However, a variety of respiratory disorders including chronic obstructive pulmonary disease or restrictive disorders such as interstitial lung disease can impair this response. 
Morbidly obese patients may also have additional restrictions on respiratory response due to their body habitus. 
On the other hand, hyperventilation in an otherwise healthy patient may make interpretation of the results difficult although the analysis of trends in other parameters such as $P_{ET_{O_2}}$ and $P{ET_{CO_2}}$ aids in the diagnosis of hyperventilation versus true respiratory problems.

\section{Role of CPET in preoperative assessment}
\label{sec:cpx_roleInAssessment}
\subsection{General surgery}

Several studies over the past 2 decades have established the value of cardiopulmonary exercise testing in patients, especially elderly, undergoing major abdominal surgery.

Older and co-workers first reported on the value of cardiopulmonary exercise testing for preoperative risk assessment in a prospective study of 187 patients over the age of 60 undergoing major abdominal surgery  \parencite{older_preoperative_1993}. 
In a later, larger prospective study of 548 patients who underwent major abdominal surgery including colorectal and abdominal aneurysm surgery, they used a risk stratification system that combined $\dot{V}_{O_2}AT<11$ml/kg/min, $\dot{V}_E/\dot{V}_{CO_2}>35$ and evidence of myocardial ischaemia during exercise. 
High risk patients ($\dot{V}_{O_2}AT<11$ml/kg/min, n=153) were admitted to intensive care after surgery and had a mortality due to cardiovascular causes of 4.6\%. 
Moderate risk group ($\dot{V}_{O_2}AT>11$ml/kg/min and either $\dot{V}_E/\dot{V}_{CO_2}>35$ or evidence of myocardial ischaemia, n=115) was managed in the high dependency unit after surgery and had a cardiovascular mortality of 1.7\%. 
Low risk patients (with none of the risk factors mentioned above, n=280) were managed on the general surgical ward with no cardiovascular mortality \parencite{older_cardiopulmonary_1999}. 
Subsequent literature reviews by Older and co-workers have emphasised the value of cardiopulmonary exercise testing in risk assessment as well as optimising perioperative care in the high-risk surgical patient. \parencite{older_preoperative_2000, older_clinical_2004, older_preoperative_2005}.
